---
title: "GNPC Data 3 Tibble Prep"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
## NOTE:
## This rmd script can only be executed in AD Workbench environment with approved access to GNPC Harmony Dataset V1.
## The input/output file paths are set as the defaulted path when the GNPC dataset V1 is approved to be imported into the AD workbench Rstudio environment.

knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(tidyverse)
library(readr)
library(ggplot2)
library(tidyr)
library(ggrepel)
library(purrr)
```

```{r, include=TRUE}
output_version <- "v6"
```

```{r, include=TRUE, echo=FALSE}
directories <- c(
  "/home/rstudio/data_prep",
  "/home/rstudio/data_prep/annotation",
  "/home/rstudio/data_prep/metadata",
  "/home/rstudio/data_prep/proteomics"
)
for (directory_path in directories) {
  dir.create(directory_path, showWarnings = FALSE, recursive = TRUE)
}
```

## 1. ANALYTE METADATA (protein target info)
```{r, include=TRUE}
analyte_tbl <- read_csv("/home/rstudio/GNPC Harmonised Dataset v1/Clinical/SomalogicAnalyteInfoV1_anonymized.csv",
  show_col_types = FALSE
)

analyte_tbl <- as_tibble(analyte_tbl |> select(column_name, everything()), index = column_name)

## Analytes contain non-human targets
unique(analyte_tbl$organism)
```


```{r, include=TRUE}
## Filter for human protein analytes
human_prot_analyte_tbl <- analyte_tbl |>
  filter(organism == "Human") |>
  filter(type == "Protein")
human_prot_analyte_columns <- human_prot_analyte_tbl$column_name

saveRDS(human_prot_analyte_tbl, paste0("/home/rstudio/data_prep/annotation/annotation_tbl_", output_version, ".rds"))

# This file is already tidy: one row per column_name or seq_id
# So no transformation needed.
```


## 2. PROTEOMICS DATA (protein measurement tibble)

```{r, include=TRUE}
proteomic_files <- list.files(
  path = "/home/rstudio/GNPC Harmonised Dataset v1/Clinical",
  pattern = "Somalogic\\d+V1_anonymized\\.csv$",
  full.names = TRUE
)

# Predefine the columns you actually need
needed_cols <- c("sample_id", "contributor_code", "visit", "sample_type")

# Function to read and clean one proteomic file at a time
read_and_clean <- function(file, keep_cols = needed_cols) {
  read_csv(
    file,
    show_col_types = FALSE,
    progress = FALSE,
    col_select = all_of(keep_cols) | starts_with("seq_")
  ) |>
    filter(sample_type == "Sample")
}

# Process the first file (for metadata)
first_tbl <- read_and_clean(proteomic_files[[1]])
metadata_tbl <- first_tbl |> select(all_of(needed_cols))

# Initialize proteomic_tbl with protein data from the first file
proteomic_tbl <- first_tbl |>
  select(sample_id, starts_with("seq_"))

# Iteratively join other files, one at a time
for (f in proteomic_files[-1]) {
  tmp_tbl <- read_and_clean(f) |>
    select(sample_id, starts_with("seq_"))
  proteomic_tbl <- full_join(proteomic_tbl, tmp_tbl, by = "sample_id")
}

# Finally, filter down to human analytes
proteomic_tbl <- proteomic_tbl |>
  select(sample_id, all_of(human_prot_analyte_columns))

# Ensure sample_id is not duplicated
proteomic_tbl <- proteomic_tbl |>
  distinct(sample_id, .keep_all = TRUE)

# There are -1 (or rather negative) values to be casted to NA
proteomic_tbl <- proteomic_tbl |>
  mutate(across(where(is.numeric), ~ ifelse(. < 0, NA, .)))

# Summarize missingness per column
missing_proteomic_na_df <- proteomic_tbl |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing_count")

# Plot histogram of missing values
ggplot(missing_proteomic_na_df, aes(x = missing_count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    title = "Histogram of Missing Values per Protein Target",
    x = "Count of Missing Values",
    y = "Count of Protein Targets"
  ) +
  theme_minimal()
```
There is a few protein targets that have high number of missingness (see the short bars on the right). But overall, there are not much missingness. The per contributor missingness will be examined later.


```{r, include=TRUE}
# proteomic_tbl now has columns as :
# sample_id | protein columns...
# and with only "Sample" values as in the original 'sample_type' col

# save to RDS for now
proteomic_tbl <- as_tibble(proteomic_tbl, index = sample_id)
saveRDS(proteomic_tbl, paste0("/home/rstudio/data_prep/proteomics/proteomic_tbl_", output_version, ".rds"))
```



## 3. CLINICAL DATA (clinical tibble with patient mapping, and sample_matrix from meta table)
```{r, include=TRUE}
# Read clinical file, person mapping file, and
clinical_raw <- read_csv("/home/rstudio/GNPC Harmonised Dataset v1/Clinical/ClinicalV1_anonymized.csv", show_col_types = FALSE)
person_map <- read_csv("/home/rstudio/GNPC Harmonised Dataset v1/Clinical/PersonMappingV1_anonymized.csv",
  show_col_types = FALSE
)
metadata_tbl <- read_csv("/home/rstudio/GNPC Harmonised Dataset v1/Clinical/SomalogicMetaV1_anonymized.csv",
  show_col_types = FALSE
)
# person_map has: person_id, sample_id
# Read lookup table
gnpc_lookup_map <- read_csv("/home/rstudio/GNPC Harmonised Dataset v1/Clinical/GNPCLookupV1_anonymized.csv",
  show_col_types = FALSE
)


# Merge metadata from protein table with clinical
clinical_plus <- metadata_tbl |>
  select(sample_id, visit, sample_type) |>
  left_join(clinical_raw, by = "sample_id")

# Merge sample matrix col from meta table with clinical
clinical_plus_2 <- metadata_tbl |>
  select(sample_id, sample_matrix) |>
  left_join(clinical_plus, by = "sample_id")

# Merge clinical + mapping
clinical_tbl <- clinical_plus_2 |>
  left_join(person_map, by = "sample_id") |>
  relocate(person_id, .after = sample_id)

# clinical_tbl now has patient IDs (person_id) alongside sample-level clinical data

# Factorize contributor
clinical_tbl <- clinical_tbl |> mutate(contributor_code = factor(contributor_code))
```

### Examine the pd changes in patients
```{r, include=TRUE}
clinical_samples_with_pd_changes <- clinical_tbl |>
  group_by(person_id) |>
  filter(n_distinct(pd) > 1) |>
  ungroup() |>
  select(sample_id, person_id, contributor_code, visit, pd) |>
  arrange(person_id, visit)


plot_pd_by_person <- function(data, jitter_width = 0.1) {
  # Check required columns
  required_cols <- c("person_id", "contributor_code", "visit", "pd")
  if (!all(required_cols %in% colnames(data))) {
    stop("Data must have columns: person_id, contributor_code, visit, pd")
  }

  data |>
    arrange(person_id, visit) |>
    ggplot(aes(
      x = visit,
      y = pd,
      group = person_id,
      color = contributor_code
    )) +
    geom_line(
      linewidth = 1,
      alpha = 0.7,
      position = position_jitter(width = jitter_width, height = 0)
    ) +
    geom_point(
      size = 3,
      position = position_jitter(width = jitter_width, height = 0)
    ) +
    scale_y_continuous(breaks = seq(-1, 2, 1)) +
    labs(
      title = "PD Diagnosis Progression by Person",
      x = "Visit",
      y = "PD Diagnosis",
      color = "Contributor"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

plot_pd_by_person(clinical_samples_with_pd_changes)
```
It appears that only 4-5 patients may have different pd labelings through a history of visits but most patients for the four important contributors (C, F, J, L, Q) have only one pd labelings (0 or 1). Conclusion: if we are OK to lose a few samples/persons from contributor J, we can accept to cast pd into just 0, 1 factors since most persons are labeled with one diagnosis of pd.


```{r, include=TRUE}
## Find out about the pd and # of samples per person distributions across the CFJLQ cohorts
clinical_tbl |>
  select(contributor_code, person_id, sample_id, pd) |>
  group_by(contributor_code, person_id, pd) |>
  summarise(n_samples = n(), .groups = "keep") |>
  group_by(contributor_code, n_samples, pd) |>
  summarise(count = n(), .groups = "keep") |>
  filter(contributor_code %in% c("C", "F", "J", "L", "Q"))
```


### Check out the proteomics (downsampled) value distributions per contributor
```{r, include=TRUE}
proteomic_tbl_w_contributor <- inner_join(clinical_tbl[c("sample_id", "contributor_code")], proteomic_tbl, by = "sample_id")

plot_contributor_median <- function(data, n_samples = 5000, seed = 123) {
  set.seed(seed)

  # Identify numeric columns
  numeric_cols <- names(data)[!(names(data) %in% c("sample_id", "contributor_code"))]

  # Optional: downsample rows for speed
  if (n_samples < nrow(data)) {
    data <- data |> slice_sample(n = n_samples)
  }

  # Compute median per sample across all numeric columns
  data <- data |>
    rowwise() |>
    mutate(sample_median = median(c_across(all_of(numeric_cols)), na.rm = TRUE)) |>
    ungroup()

  # Summarize per contributor
  summary_df <- data |>
    group_by(contributor_code) |>
    summarise(
      median_of_medians = median(sample_median, na.rm = TRUE),
      q25 = quantile(sample_median, 0.25, na.rm = TRUE),
      q75 = quantile(sample_median, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  # Plot: median with IQR as error bars
  ggplot(summary_df, aes(x = contributor_code, y = median_of_medians)) +
    geom_point(size = 3, color = "blue") +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2, color = "darkgray") +
    theme_minimal() +
    labs(
      title = "Sample Median of Measurements by Contributor",
      x = "Contributor",
      y = "Median of Sample Medians"
    )
}
plot_contributor_median(proteomic_tbl_w_contributor, n_samples = 1000)
```

### Check out the proteomics (downsampled) value missingness per contributor
```{r, include=TRUE}
plot_contributor_missingness <- function(data, n_samples = 5000, seed = 123) {
  set.seed(seed)

  # Optional: downsample rows for speed
  if (n_samples < nrow(data)) {
    data <- data |> slice_sample(n = n_samples)
  }

  # Identify numeric/measurement columns
  measurement_cols <- names(data)[!(names(data) %in% c("sample_id", "contributor_code"))]

  # Compute fraction of missing values per sample
  data <- data |>
    rowwise() |>
    mutate(missing_frac = sum(is.na(c_across(all_of(measurement_cols)))) / length(measurement_cols)) |>
    ungroup()

  # Summarize per contributor
  summary_df <- data |>
    group_by(contributor_code) |>
    summarise(
      mean_missing = mean(missing_frac, na.rm = TRUE),
      median_missing = median(missing_frac, na.rm = TRUE),
      q25 = quantile(missing_frac, 0.25, na.rm = TRUE),
      q75 = quantile(missing_frac, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  # Plot
  ggplot(summary_df, aes(x = contributor_code, y = median_missing)) +
    geom_point(size = 3, color = "red") +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2, color = "darkgray") +
    theme_minimal() +
    labs(
      title = "Missingness per Contributor",
      x = "Contributor",
      y = "Median Fraction of Missing Values"
    )
}

plot_contributor_missingness(proteomic_tbl_w_contributor, n_samples = 1000)
```

### Alternatively check out the proteomics (downsampled) value distribution using density plot per contributor
```{r, include=TRUE}
plot_contributor_density <- function(data, n_samples = 5000, n_features = 200,
                                     lower_pct = 0.001, upper_pct = 0.999,
                                     seed = 123) {
  set.seed(seed)

  # Downsample rows
  if (n_samples < nrow(data)) {
    data <- data |> slice_sample(n = n_samples)
  }

  # Select numeric measurement columns
  numeric_cols <- setdiff(names(data), c("sample_id", "contributor_code"))

  # Downsample columns
  if (n_features < length(numeric_cols)) {
    numeric_cols <- sample(numeric_cols, n_features)
  }

  # Convert to long format
  long_df <- data |>
    select(contributor_code, all_of(numeric_cols)) |>
    pivot_longer(
      cols = all_of(numeric_cols),
      names_to = "feature",
      values_to = "value"
    ) |>
    filter(!is.na(value))

  # Percentile capping
  q_low <- quantile(long_df$value, lower_pct, na.rm = TRUE)
  q_high <- quantile(long_df$value, upper_pct, na.rm = TRUE)

  long_df <- long_df |>
    mutate(value = pmin(pmax(value, q_low), q_high))

  # Calculate median for label positions
  label_positions <- long_df |>
    group_by(contributor_code) |>
    summarise(median_val = median(value, na.rm = TRUE), .groups = "drop")

  # Density plot
  ggplot(long_df, aes(x = value, color = contributor_code, fill = contributor_code)) +
    geom_density(alpha = 0.3) +
    scale_x_log10() +
    geom_text_repel(
      data = label_positions,
      aes(x = median_val, y = 0, label = contributor_code, color = contributor_code),
      inherit.aes = FALSE,
      size = 3,
      direction = "y",
      nudge_y = 0.05,
      show.legend = FALSE
    ) +
    theme_minimal() +
    labs(
      title = paste("Density of Measurements per Contributor (log-scale, capped at ",
        lower_pct * 100, "–", upper_pct * 100, " percentiles)",
        sep = ""
      ),
      x = "Measurement Value (log10 scale)",
      y = "Density",
      color = "Contributor",
      fill = "Contributor"
    )
}

plot_contributor_density(proteomic_tbl_w_contributor,
  n_samples = 1000, n_features = 2000
)
```

#### Clean up clinical table
```{r, include=TRUE}
# Function to label NA values in clinical table
clean_clinical_tbl <- function(df) {
  # Lookup table: columns mapped to values to convert to NA
  na_codes <- list(
    age_at_visit = c(-1),
    sex = c(-1),
    race = c(-1),
    bmi = c(-1),
    years_of_education = c(-1),
    computed_age_range = c(-1),
    computed_years_education_range = c(-1),
    computed_height_range = c(-2, -1),
    computed_weight_range = c(-1),
    computed_bmi_range = c(-1),
    computed_pulse_range = c(-1),
    computed_systolic_blood_pressure_range = c(-1),
    computed_diastolic_blood_pressure_range = c(-2, -1),
    alcohol_hx = c(-1),
    smoking_hx = c(-1),
    computed_years_smoked = c(-2, -1),
    total_years_smoked = c(-1),
    stroke = c(-1, 2), # 2 = Unsure
    tia = c(-1, 2),
    tbi = c(-1, 2),
    recruited_control = c(-1, 2),
    ad = c(-1, 2),
    ftd = c(-1, 2),
    pd = c(-1, 2),
    als = c(-1, 2),
    mci_sci = c(-1, 2),
    cancer = c(-1, 2),
    diabetes = c(-1, 2),
    chf = c(-1, 2),
    copd = c(-1, 2),
    mi = c(-1, 2),
    afib = c(-1, 2),
    angina = c(-1, 2),
    hyperlipidaemia = c(-1, 2),
    hypertension = c(-1, 2),
    depression = c(-1, 2),
    depression_test_battery = c("N/A"),
    anxiety = c(-1),
    cdr = c(-1),
    computed_cognitive_test_score = c(-6, -5, -4, -3, -2, -1), ## all neg scores mean incomplete
    cognitive_test_battery = c("N/A"),
    computed_clinical_diagnosis = c(-1),
    computed_cognitive_impairment = c(-1),
    is_neuropath = c(-1),
    is_biomarker = c(-1),
    cognitive_test_date = c("1/1/1900"),
    resting_heart_rate_pulse = c(-1),
    systolic_blood_pressure_sitting = c(-1),
    diastolic_blood_pressure_sitting = c(-1),
    depression_inventory_score = c(-1),
    cognitive_test_score = c(-1),
    sample_matrix = c("-1")
  )

  # Loop over each column and replace codes with NA if applicable
  for (col in names(na_codes)) {
    if (col %in% names(df)) {
      df[[col]] <- ifelse(df[[col]] %in% na_codes[[col]], NA, df[[col]])
    }
  }

  return(df)
}

# clean up clinical_tbl
cleaned_clinical_tbl <- clean_clinical_tbl(clinical_tbl)

map_clinical_tbl <- function(df, lookup) {
  # Simplify descriptions (make concise but meaningful)
  simplify_desc <- function(desc) {
    desc <- trimws(desc)
    desc <- gsub("Not Stated", NA, desc)
    desc <- gsub("N/A", NA, desc)
    desc <- gsub("Z=-1 to Z=1", "Normal abs(Z) <= 1", desc)
    desc <- gsub("Z=-2 to Z=-1 and Z=1 to Z=2", "Mild dev abs(Z) <= 2", desc)
    desc <- gsub("Z=-3 to Z=-2 and Z=2 to Z=3", "Moderate dev abs(Z) <= 3", desc)
    desc <- gsub("Z<3 and Z>3", "Severe dev abs(Z) > 3", desc)
    desc <- gsub("Incompatible data", "Invalid", desc)
    desc <- gsub("Yes", "Yes", desc)
    desc <- gsub("No", "No", desc)
    desc <- gsub("Unsure", NA, desc)
    desc <- gsub("Male", "Male", desc)
    desc <- gsub("Female", "Female", desc)
    desc <- gsub("Never_or_Previous", "Never/Prev", desc)
    desc <- gsub("Current", "Current", desc)
    desc <- gsub("Previous", "Previous", desc)
    desc <- gsub("Other\\(.*\\)", "Other", desc)
    desc <- gsub("_", " ", desc)
    desc
  }

  lookup$key_description <- vapply(lookup$key_description, simplify_desc, FUN.VALUE = character(1))

  # Build nested mapping list
  map_list <- lookup |>
    filter(!is.na(key_description)) |>
    group_by(column_name) |>
    summarize(mapping = list(setNames(key_description, key_lookup)), .groups = "drop")

  map_dict <- setNames(map_list$mapping, map_list$column_name)

  # Apply mappings
  for (col in names(df)) {
    if (col %in% names(map_dict)) {
      # Replace using named vector
      df[[col]] <- as.character(df[[col]])
      df[[col]] <- recode(df[[col]], !!!map_dict[[col]])
      df[[col]][df[[col]] == ""] <- NA # empty strings → NA
    }
  }

  return(df)
}

# map clinical_tbl
mapped_cleaned_clinical_tbl <- map_clinical_tbl(cleaned_clinical_tbl, gnpc_lookup_map)

# Process BMI
mapped_cleaned_clinical_tbl <- mapped_cleaned_clinical_tbl |>
  mutate(
    bmi_clean = str_replace_all(bmi, "[^0-9\\.-]", ""),
    bmi_low = as.numeric(str_extract(bmi_clean, "^[0-9]+\\.?[0-9]*")),
    bmi_high = as.numeric(str_extract(bmi_clean, "(?<=-)[0-9]+\\.?[0-9]*")),
    bmi_mid = ifelse(!is.na(bmi_high), (bmi_low + bmi_high) / 2, bmi_low),
    bmi_cat = case_when(
      is.na(bmi_mid) ~ NA_character_,
      bmi_mid < 18.5 ~ "underweight",
      bmi_mid < 25 ~ "normal",
      bmi_mid < 30 ~ "overweight",
      bmi_mid < 35 ~ "obesity_1",
      bmi_mid < 40 ~ "obesity_2",
      TRUE ~ "obesity_3"
    )
  )


# Numeric coercions
num_cols <- c(
  "years_of_education", "age_at_visit", "computed_years_education_range",
  "computed_height_range", "computed_weight_range", "computed_bmi_range",
  "resting_heart_rate_pulse", "computed_pulse_range", "systolic_blood_pressure_sitting",
  "diastolic_blood_pressure_sitting", "total_years_smoked", "depression_inventory_score",
  "cognitive_test_score", "bmi_low", "bmi_high", "bmi_mid"
)
mapped_cleaned_clinical_tbl <- mapped_cleaned_clinical_tbl |> mutate(across(all_of(num_cols), ~ as.numeric(.)))

# categorical coercions
cat_cols <- c(
  "computed_age_range", "sex", "race", "computed_years_education_range",
  "computed_height_range", "computed_weight_range", "computed_bmi_range",
  "computed_pulse_range", "computed_systolic_blood_pressure_range",
  "computed_diastolic_blood_pressure_range", "alcohol_hx", "smoking_hx",
  "computed_years_smoked", "stroke", "tia", "tbi", "recruited_control",
  "ad", "ftd", "pd", "als", "mci_sci", "cancer", "diabetes", "chf", "copd",
  "mi", "afib", "angina", "hyperlipidaemia", "hypertension", "depression",
  "anxiety", "cdr", "computed_cognitive_test_score", "computed_clinical_diagnosis",
  "computed_cognitive_impairment", "is_neuropath", "is_biomarker", "is_somalogic",
  "is_mass_spec", "bmi_clean", "depression_test_battery", "cognitive_test_date",
  "cognitive_test_battery", "bmi", "bmi_cat"
)
mapped_cleaned_clinical_tbl <- mapped_cleaned_clinical_tbl |> mutate(across(all_of(cat_cols), ~ as.factor(.)))

# save to RDS for now
mapped_cleaned_clinical_tbl <- as_tibble(mapped_cleaned_clinical_tbl, index = sample_id)
saveRDS(mapped_cleaned_clinical_tbl, paste0("/home/rstudio/data_prep/metadata/clinical_metadata_tbl_", output_version, ".rds"))
```


## 4. Save per Cohort Data (select individual contributor CFJLQ) 
```{r, include=TRUE}
# selected cohorts (contributors)
selected_contributors <- c("C", "F", "J", "L", "Q")

for (sel_con in selected_contributors) {
  clin_rds_file_per_cohort <- paste0("/home/rstudio/data_prep/metadata/clinical_metadata_tbl_", output_version, "_contributor_", sel_con, ".rds")
  clin_tbl_per_cohort <- mapped_cleaned_clinical_tbl |>
    filter(contributor_code == sel_con)
  saveRDS(clin_tbl_per_cohort, clin_rds_file_per_cohort)

  sample_ids_per_cohort <- clin_tbl_per_cohort$sample_id
  cat("Contributor Code", sel_con, length(sample_ids_per_cohort), "Samples\n")

  prot_rds_file_per_cohort <- paste0("/home/rstudio/data_prep/proteomics/proteomic_tbl_", output_version, "_contributor_", sel_con, ".rds")
  proteomic_tbl_per_cohort <- proteomic_tbl |>
    filter(sample_id %in% sample_ids_per_cohort)
  saveRDS(proteomic_tbl_per_cohort, prot_rds_file_per_cohort)
}
```
