---
title: "UKB-RAP: Protein Expression Analysis"
subtitle: "Single Protein and Protein Ratio"
author:
  - name: "Nathan Watson-Haigh"
date-modified: last-modified
date-format: "YYYY-MM-DDTHH:mm:ssZ"
format:
  html:
    embed-resources: true
    self-contained: true
editor_options:
  chunk_output_type: console
toc: true
toc-depth: 2
---

## Setup

```{r}
#| label: setup

here::i_am("src/20_analyses/target_validation/uk_biobank/index.qmd")

ukb_rap_working_dir <- "Users/Nathan"

assets_dir <- here::here("assets")
functions_dir <- here::here("functions")

#####
# Set which files to use as input to this quarto document
#####
case_control_pairs <- glue::glue("{ukb_rap_working_dir}/cohorts/pd_case_control_pairs.new.tsv.gz")
pair_files <- c(
  list_1 = here::here(glue::glue("{staging_input_data_dir}/list_1_v29452.tsv.gz")),
  list_2 = here::here(glue::glue("{staging_input_data_dir}/list_2_v29451.tsv.gz")),
  list_3 = here::here(glue::glue("{staging_input_data_dir}/list_3_v29453.tsv.gz"))
)
window_specs <- tibble::tribble(
  ~"window_id",      ~"window_start", ~"window_end",
  "[-2,0)",          -2,               0,
  "[-3,-1)",         -3,              -1,
  "[-4,-2)",         -4,              -2,
  "[-5,-3)",         -5,              -3,
  "[-6,-4)",         -6,              -4,
  "[-7,-5)",         -7,              -5,
  "[-8,-6)",         -8,              -6,
  "[-10,-8)",       -10,              -8,
  "[-Inf,-1)",     -Inf,              -1,
  "[-Inf,-7)",     -Inf,              -7,
  "[0,Inf)",          0,             Inf,
  "[-9,-5)",         -9,              -5,
  "[-7,-3)",         -7,              -3,
  "[-5,-1)",         -5,              -1,
  "[0,2)",            0,               2,
  "[1,3)",            1,               3,
  "[2,4)",            2,               4,
  "[3,5)",            3,               5,
)
#####

timestamp <- lubridate::now() |> format("%Y%m%d_%H%M%S_%z")

staging_parent_dir <- here::here(glue::glue("staging/protein_ratio_analysis/{timestamp}"))
staging_input_data_dir <- here::here(staging_parent_dir, "inputs")
staging_output_data_dir <- here::here(staging_parent_dir, "outputs", "data")
staging_output_plot_dir <- here::here(staging_parent_dir, "outputs", "plots")
purrr::walk(
  c(staging_input_data_dir, staging_output_plot_dir, staging_output_plot_dir, staging_output_data_dir),
  dir.create,
  showWarnings = FALSE,
  recursive = TRUE,
)

all_results <- list()
cobalt_long <- list()
cobalt_plot <- list()
cobalt_hist_plots_pdf <- list()
volcano_plot <- list()

.transform_data <- function(.data) {
  .data |>
    tidyr::pivot_longer(
      cols = c(
        tidyselect::starts_with("auc_assessment_mean"),
        tidyselect::starts_with("sensitivity"),
        tidyselect::starts_with("specificity"),
        tidyselect::starts_with("accuracy")
      ),
      names_to = c("measure", "interval_start", "interval_end"),
      names_sep = "_(?=-?(\\d+|Inf))",
      values_to = "value"
    )
}
```

### Stage Input Data

#### Protein Pairs of Interest (PPOI)

```{r}
#| label: stage-protein-pairs-of-interest

dir.create(here::here("assets"), recursive = TRUE, showWarnings = FALSE)

glue::glue(,
  "{get_config()$input_dir}/name={name}/version={version}/dataset.parquet",
  name = "Chronos_phase1_protein_ratio_list_1",
  version = 29452
) |>
    arrow::open_dataset() |>
    dplyr::collect() |>
    .transform_data() |>
    readr::write_tsv(
        pair_files[["list_1"]]
    )

glue::glue(,
  "{get_config()$input_dir}/name={name}/version={version}/dataset.parquet",
  name = "Chronos_phase1_protein_ratio_list_1",
  version = 29452
) |>
    arrow::open_dataset() |>
    dplyr::collect() |>
    .transform_data() |>
    readr::write_tsv(
        pair_files[["list_2"]]
    )

glue::glue(,
  "{get_config()$input_dir}/name={name}/version={version}/dataset.parquet",
  name = "Chronos_phase1_protein_ratio_list_3",
  version = 29453
) |>
    arrow::open_dataset() |>
    dplyr::collect() |>
    .transform_data() |>
    readr::write_tsv(
        pair_files[["list_3"]]
    )
```

#### Case/Control Pairings

```{r}
#| label: stage-case-control-pairings

case_control_pairing_file <- here::here(staging_input_data_dir, "pd_case_control_pairs.tsv.gz")

ukb_rap_download(
  rap_path = case_control_pairs,
  local_path = case_control_pairing_file
)
```

#### Assay ID to Gene Symbol Mappings

```{r}
#| label: stage-assay_id2gene_symbol

#####
# Generate Lookup: assay_id_2_gene_symbol
#####
assay_id2gene_symbol_mapping_file <- here::here(glue::glue("{staging_input_data_dir}/protein_platform_uniprot_ids_v29404.tsv.gz"))

glue::glue("{get_config()$input_dir}/name=protein_platform_uniprot_ids/version=29404/dataset.parquet") |>
  arrow::open_dataset() |>
  dplyr::collect() |>
  readr::write_tsv(
    assay_id2gene_symbol_mapping_file
  )
```

### Load Staged Data

```{r}
#| label: load-protein-pairs-of-interest

ppoi <- readr::read_tsv(
  pair_files,
  show_col_types = FALSE,
  col_types = readr::cols_only(
    protein1 = readr::col_character(),
    protein2 = readr::col_character(),
  ),
  id = "File"
) |>
  dplyr::distinct() |>
  dplyr::rename(
    assay_id_numerator = protein1,
    assay_id_denominator = protein2,
  ) |>
  dplyr::mutate(
    pair_id = stringr::str_replace_all(glue::glue("{assay_id_numerator}_o_{assay_id_denominator}"), "-", "_")
  )
```

```{r}
#| label: load-case-control-pairings

case_control_pairs <- readr::read_tsv(
  case_control_pairing_file,
  show_col_types = FALSE
)
```

```{r}
#| label: load-assay_id2gene_symbol

assay_id_2_gene_symbol <- readr::read_tsv(
  assay_id2gene_symbol_mapping_file,
  col_types = readr::cols_only(
    platform_id = readr::col_factor(),
    gene_symbol = readr::col_factor(),
  )
) |>
  dplyr::rename(
     assay_id = platform_id
  ) |>
  tidyr::separate_rows(
    gene_symbol,
    sep = ":"
  ) |>
  dplyr::distinct(
    assay_id,
    gene_symbol
  ) |>
  tidyr::drop_na()
```

```{r}
#| label: report-assy-id-2-gene-symbols-multiplicity

# NOTE: Some of the `assay_id`s have multiple `gene_symbol`s associated with them.
assay_id_2_gene_symbol |>
  dplyr::group_by(assay_id) |>
  dplyr::filter(
    dplyr::n() > 1
  )
```

```{r}
#| label: map-gene-symbols-to-ppoi

ppoi <- ppoi |>
  #####
  # Add gene symbols for numerators so we can map between our PPOIs and UKB Data
  #####
  dplyr::left_join(
    assay_id_2_gene_symbol |>
      dplyr::rename(
        assay_id_numerator = assay_id,
        gene_symbol_numerator = gene_symbol,
      ),
    by = dplyr::join_by(assay_id_numerator),
    relationship = "many-to-many"
  ) |>
  #####
  # Add gene symbols for denominators so we can map between our PPOIs and UKB Data
  #####
  dplyr::left_join(
    assay_id_2_gene_symbol |>
      dplyr::rename(
        assay_id_denominator = assay_id,
        gene_symbol_denominator = gene_symbol,
      ),
    by = dplyr::join_by(assay_id_denominator),
    relationship = "many-to-many"
  )
```

### Define Data Codings Used by Various UKB Data Fields

```{r}
#| label: load-ukb-data-coding

data_coding <- list()

# https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=9
data_coding[[9L]] <- tibble::tribble(
  ~"coding",  ~"meaning",
  0L,         "Female",
  1L,         "Male",
)
```

### Get All UKB Participants

```{r}
#| label: load-all-ukb-participants

all_participants <- tibble::tribble(
  ~"field",          ~"label",
  "participant.eid", "Participant_ID",
) |>
  dplyr::filter(
    field == "participant.eid"
  ) |>
  ukb_rap_get_cohort_data(
    cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
  )
```

#### Get/Define Participant Metadata:

##### Common Metadata

```{r}
#| label: load-participant-metadata

#####
# Year of Birth
#####
all_participants <- all_participants |>
  dplyr::left_join(
    tibble::tribble(
      ~"field",                  ~"label",
      "participant.eid",         "Participant_ID",
      "participant.p34",         "Year of Birth",
    ) |>
      ukb_rap_get_cohort_data(
        cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
      ) |>
      dplyr::filter(
        !is.na(coding)
      ) |>
      dplyr::select(
        Participant_ID,
        YOB = coding
      ),
    by = dplyr::join_by("Participant_ID")
  )

#####
# Date of Death
#####
all_participants <- all_participants |>
  dplyr::left_join(
    tibble::tribble(
      ~"field",                  ~"label",
      "participant.eid",         "Participant_ID",
      "participant.p40000_i0",   "Date of Death | Instance 0",
      "participant.p40000_i1",   "Date of Death | Instance 1",
    ) |>
      ukb_rap_get_cohort_data(
        cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
      ) |>
      dplyr::filter(
        !is.na(coding)
      ) |>
      dplyr::distinct(
        Participant_ID,
        DOD = coding
      ),
    by = dplyr::join_by("Participant_ID")
  )

#####
# Sex
#####
all_participants <- all_participants |>
  dplyr::left_join(
    tibble::tribble(
      ~"field",                  ~"label",
      "participant.eid",         "Participant_ID",
      "participant.p31",         "Sex",
    ) |>
      ukb_rap_get_cohort_data(
        cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
      ) |>
      dplyr::filter(
        !is.na(coding)
      ) |>
      #####
      # Decode the Sex field
      #####
      dplyr::left_join(
        data_coding[[9L]],
        by = dplyr::join_by("coding")
      ) |>
      dplyr::select(
        Participant_ID,
        SEX = meaning
      ),
    by = dplyr::join_by("Participant_ID")
  )

#####
# Townsend Deprivation Index (TDI)
#####
all_participants <- all_participants |>
  dplyr::left_join(
    tibble::tribble(
      ~"field",                  ~"label",
      "participant.eid",         "Participant_ID",
      "participant.p22189",      "Townsend deprivation index at recruitment",
    ) |>
      ukb_rap_get_cohort_data(
        cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
      ) |>
      dplyr::filter(
        !is.na(coding)
      ) |>
      dplyr::select(
        Participant_ID,
        TDI = coding
      ),
    by = dplyr::join_by("Participant_ID")
  )
```

##### Disease-Specific Metadata

UKB data fields:

```{r}
#| label: load-disease-specific-participant-metadata

#####
# PD Onset Date: Date G20 first reported
#####
all_participants <- all_participants |>
  dplyr::left_join(
    tibble::tribble(
      ~"field",                  ~"label",
      "participant.eid",         "Participant_ID",
      "participant.p131022",     "Date G20 first reported (parkinson's disease)",
    ) |>
      ukb_rap_get_cohort_data(
        cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
      ) |>
      dplyr::filter(
        !is.na(coding)
      ) |>
      dplyr::select(
        Participant_ID,
        ONSET_DATE.PD = coding
      ),
    by = dplyr::join_by("Participant_ID")
  )
```

Derived data fields:

```{r}
#| label: calc-disease-specific-participant-metadata

all_participants <- all_participants |>
  dplyr::mutate(
    AGE_OF_ONSET.PD = lubridate::interval(lubridate::make_date(year = YOB), ONSET_DATE.PD) / lubridate::years(1),
  )
```

### Load Disease Status

```{r}
#| label: load-cohort-disease-statuses

disease_status <- all_participants |>
  #####
  # Get the field 41270 data
  #####
  dplyr::left_join(
    tibble::tribble(
      ~"field",               ~"label",
      "participant.eid",      "Participant_ID",
      "participant.p41270",   "Diagnoses - ICD10",
    ) |>
    ukb_rap_get_cohort_data(
      cohort_rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/All Participants")
    ),
    by = dplyr::join_by("Participant_ID")
  ) |>
  #####
  # Parse out the ICD10 values
  #####
  dplyr::select(
    Participant_ID,
    p41270 = coding
  ) |>
  dplyr::mutate(
    p41270 = stringr::str_remove_all(p41270, "\\[|\\]|\"")
  ) |>
  tidyr::separate_rows(
    p41270,
    sep = ","
  ) |>
  #####
  # Determine who has what diseases based on the ICD10 codes found
  #####
  dplyr::group_by(Participant_ID) |>
  dplyr::summarise(
    HAS_HYPERTENSION = any(stringr::str_starts(.data[["p41270"]], "I10")),
    HAS_T2D =          any(stringr::str_starts(.data[["p41270"]], "E11")),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    dplyr::across(where(is.logical), ~ tidyr::replace_na(.x, FALSE))
  )
```

### Load Proteomics Data

```{r}
#| label: load-proteomics-data

proteomics_data <- ukb_rap_load_proteomics_data(rap_path = glue::glue("{ukb_rap_working_dir}/cohorts/With Proteomics.proteomics.RData"))

# Ignore he stored participant metadata, we'll use the metadata already obtained above from UKB
proteomics_data[["participant_metadata"]] <- NULL
proteomics_data[["row_metadata"]] <- proteomics_data[["row_metadata"]] |>
  dplyr::rename(
    Participant_ID = `Participant ID`
  )

# Add col metadata
proteomics_data[["col_metadata"]] <- tibble::tibble(
  gene_symbol = proteomics_data[["data"]] |>
    dplyr::select(-Sample_ID) |>
    colnames()
)
```

#### Restrict Proteomics Data to That Required for the Analysis

Subset the data to the participants of interest:

```{r}

#####
# participants - rows
#####
required_rows <- proteomics_data[["row_metadata"]] |>
  dplyr::semi_join(
    case_control_pairs,
    by = dplyr::join_by(Participant_ID)
  ) |>
  dplyr::select(Sample_ID)

#####
# Subset the metadata first
#####
proteomics_data[["row_metadata"]] <- proteomics_data[["row_metadata"]] |>
  dplyr::semi_join(
    required_rows,
    by = dplyr::join_by(Sample_ID)
  )

#####
# Subset the data second
#####
proteomics_data[["data"]] <- proteomics_data[["data"]] |>
  #####
  # Select the samples
  #####
  dplyr::semi_join(
    proteomics_data[["row_metadata"]],
    by = dplyr::join_by(Sample_ID)
  )
```

```{r}

proteomics_data_long <- proteomics_data[["data"]] |>
  # Make data long format
  tidyr::pivot_longer(
    cols = -Sample_ID,
    names_to = "Protein",
    values_to = "NPX"
  )
```

#### Define Sample-Level Metadata

##### Common Metadata

```{r}
#| label: calc-sample-metadata

proteomics_data[["row_metadata"]] <- proteomics_data[["row_metadata"]] |>
  dplyr::left_join(
    all_participants |>
      dplyr::select(
        Participant_ID,
        YOB
      ),
    by = dplyr::join_by(Participant_ID)
  ) |>
  dplyr::mutate(
    AGE = lubridate::interval(lubridate::make_date(year = YOB), `Date of attending assessment centre`) / lubridate::years(1),
  ) |>
  dplyr::select(
    -YOB
  )
```

##### Case/Control Pair Metadata

```{r}
#| label: calc-cohort-specific-sample-metadata

#####
# Case/Control Metadata
#####
proteomics_data[["row_metadata"]] <- proteomics_data[["row_metadata"]] |>
  dplyr::left_join(
    case_control_pairs,
    by = dplyr::join_by(Participant_ID)
  )

#####
# Derived fields
#####
proteomics_data[["row_metadata"]] <- proteomics_data[["row_metadata"]] |>
  #####
  # Join on ONSET_DATE.PD
  #####
  dplyr::left_join(
    all_participants |>
      dplyr::select(
        Participant_ID,
        ONSET_DATE.PD
      ),
    by = dplyr::join_by(Participant_ID)
  ) |>
  #####
  # Within each pair, fill ONSET_DATE.PD using the first value
  #####
  dplyr::group_by(pair) |> 
  dplyr::mutate(
    ONSET_DATE.PD = dplyr::first(ONSET_DATE.PD, na_rm = TRUE)
  ) |> 
  dplyr::ungroup() |>
  #####
  # Calculate other derived fields
  #####
  dplyr::mutate(
    YEARS_SINCE_ONSET.PD = lubridate::interval(ONSET_DATE.PD, `Date of attending assessment centre`) / lubridate::years(1),
  ) |>
  dplyr::select(
    -ONSET_DATE.PD
  )
```

## Proteomics Analysis

```{r}
#| label: functions-regression

gene_symbols <- list()

.save_dataset <- function(tbl, filename) {
  
  dir.create(
    dirname(filename),
    showWarnings = FALSE,
    recursive = TRUE,
  )
  
  if (stringr::str_ends(filename, "parquet")) {
    tbl |> arrow::write_parquet(filename)
  } else if (stringr::str_ends(filename, "tsv|tsv.gz")) {
    tbl |> readr::write_tsv(filename)
  } else if (stringr::str_ends(filename, "csv|csv.gz")) {
    tbl |> readr::write_csv(filename)
  } else {
    stop(glue::glue("Unsupported file suffix in filename: {filename}"))
  }
  
  invisible(tbl)
}
```

```{r}
#| label: prep-proteomics-data

gene_symbols[["single_protein"]] <- proteomics_data[["col_metadata"]]

single_protein_data <- proteomics_data_long |>
  dplyr::semi_join(
    gene_symbols[["single_protein"]],
    by = dplyr::join_by(Protein == gene_symbol)
  ) |>
  dplyr::left_join(
    proteomics_data[["row_metadata"]],
    by = dplyr::join_by("Sample_ID")
  ) |>
  dplyr::left_join(
    disease_status,
    by = dplyr::join_by(Participant_ID)
  ) |>
  #####
  # Join participant metadata
  #####
  dplyr::left_join(
    all_participants,
    by = dplyr::join_by(Participant_ID)
  ) |>
  #####
  # Make some variables a factor
  #####
  dplyr::mutate(
    group = factor(group, levels = c("control.eid", "case.eid")),
    SEX = factor(SEX),
    HAS_T2D = factor(HAS_T2D),
    HAS_HYPERTENSION = factor(HAS_HYPERTENSION),
    YEARS_SINCE_ONSET_CATEGORY = cut(
      YEARS_SINCE_ONSET.PD,
      breaks = c(-Inf, -10, -5, 0, 5, 10, Inf),
      right = FALSE
    )
  ) |>
  dplyr::filter(
    !is.na(NPX)
  )

future::plan(future::multisession)
```

### Single Protein Linear Regression Analysis

Let's start with the simplest linear regression with predictors of group, years since PD onset, and the covariates you used for logistic regression.

#### Perform Adjusted Linear Regression

```{r}

all_results[["single_protein"]] <- single_protein_data |>
  dplyr::rename(
    YEARS_SINCE_ONSET = YEARS_SINCE_ONSET.PD
  ) |>
  dplyr::group_by(
    Protein,
  ) |>
  tidyr::nest() |>
  # Set an explicit column ordering
  dplyr::select(
    Protein,
    data,
  ) |>
  furrr::future_pmap(function(Protein, data) {
    # Force future to send this global variable to each worker
    staging_dir <- staging_output_data_dir
    
    data |>
      #####
      # For participants with more than 1 sample in the window, pick the middle one
      #####
      pick_single_sample_per_participant(strategy = "middle") |>
      #####
      # Save each dataset to a HIVE-style directory structure of parquet files
      #####
      .save_dataset(
        filename = glue::glue("{staging_dir}/adjusted_linear_regression_data/protein={Protein}/lr_data.parquet")
      ) |>
      #####
      # Apply the adjusted linear regression model
      #####
      apply_linear_regression_model_to_data(
        model = stats::as.formula("NPX ~ group + YEARS_SINCE_ONSET_CATEGORY + SEX + YOB"),
        exposure = "groupcase.eid"
      ) |>
      #####
      # Add info about time-slice requested
      #####
      dplyr::mutate(
        Protein = !!Protein,
      )
    },
    .options = furrr::furrr_options(seed = TRUE)
  ) |>
  dplyr::bind_rows() |>
  #####
  # Ensure the columns are in the same order as previous results
  #####
  dplyr::relocate(
    Protein,
    n_samples,
  )
```

#### Save Results

```{r}
#| label: write-results

all_results[["single_protein"]] |>
  readr::write_tsv(
    here::here(staging_output_data_dir, "ukb_single_protein_adjusted_linear_regression.tsv.gz")
  )
```

#### Plot Results

##### Balance Measures in Covariates

```{r}

cobalt_long[["single_protein"]] <- all_results[["single_protein"]] |>
  dplyr::select(tidyselect::starts_with("cobalt_")) |>
  tidyr::pivot_longer(
    cols = tidyselect::everything(),
    names_to = "covariate",
    values_to = "value"
  ) |>
  dplyr::mutate(
    interval_colours = dplyr::case_when(
      abs(value) >= 0.2 ~ "lightcoral",
      abs(value) >= 0.1 ~ "tan1",
      .default = "palegreen3"
    )
  )

cobalt_plot[["single_protein"]] <- cobalt_long[["single_protein"]] |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = value,
      fill = interval_colours
    )
  ) +
  ggplot2::geom_histogram(
    binwidth = 0.025,
    colour = "white",
  ) +
  ggplot2::facet_wrap(
    ~ covariate,
    scales = "free_y",
    labeller = ggplot2::labeller(
      covariate = function(x) gsub("^cobalt_", "", x)
    )
  ) +
  ggplot2::scale_fill_identity() +
  ggplot2::labs(
    x = "Balance Statistic",
    y = "Count"
  ) +
  ggplot2::expand_limits(
    x = c(-1, 1)
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-1, 1, by = 0.2)
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  )

cobalt_hist_plots_pdf[["single_protein"]] <- here::here(glue::glue("{staging_output_plot_dir}/cobalt_hist_single_protein.pdf"))
dir.create(dirname(cobalt_hist_plots_pdf[["single_protein"]]), showWarnings = FALSE, recursive = TRUE)

pdf(cobalt_hist_plots_pdf[["single_protein"]], width = 15.0, height = 5.0)
cobalt_plot[["single_protein"]]
dev.off()
```

##### Volcano Plots

```{r}
#! label: prep-plot-volcano

#####
# Setup base volcano plot
#####
volcano_plot[["single_protein"]] <- all_results[["single_protein"]] |>
  plot_volcano(x = "beta", title = "Volcano Plot of Adjusted Linear Regression Results", x_lab = "beta")
```

```{r}
#! label: plot-volcanos

#####
# Plot a single volcano containing all results
#####
ggplot2::ggsave(
  filename = here::here(glue::glue("{staging_output_plot_dir}/volcano_single_protein.pdf")),
  plot = volcano_plot[["single_protein"]],
  width = 6.0, height = 4.0, units = "in",
  dpi = 600,
  bg = "white"
)
```

##### Protein NPX

```{r}
#! label: plot-pairs-expression-profiles

all_proteins_pdf <- list()

all_proteins_pdf[["single_protein"]] <- here::here(glue::glue("{staging_output_plot_dir}/all_proteins_UKB.pdf"))
dir.create(dirname(all_proteins_pdf[["single_protein"]]), showWarnings = FALSE, recursive = TRUE)

subdir_gene_symbol <- here::here(glue::glue("{staging_output_plot_dir}/proteins"))
dir.create(subdir_gene_symbol, showWarnings = FALSE, recursive = TRUE)
    
single_protein_data |>
  dplyr::rename(
    YEARS_SINCE_ONSET = YEARS_SINCE_ONSET.PD
  ) |>
  dplyr::group_by(Protein) |>
  tidyr::nest() |>
  purrr::pwalk(function(Protein, data) {
    
    p <- data |>
      plot_protein(y_var = "NPX", symmetric.y = TRUE) +
      ggplot2::facet_wrap(
        glue::glue("{Protein}<br /><span style='font-size:7pt;color:gray;'>{Protein}</span>") ~ .,
        ncol = 1L,
        scales = "free_y"
      ) +
      ggplot2::theme(
        strip.text.x = ggtext::element_markdown()
      )
    
    # Also create individual plot files
    protein_plot_file <- here::here(glue::glue("{staging_output_plot_dir}/proteins/{Protein}_UKB.pdf"))
    dir.create(dirname(protein_plot_file), showWarnings = FALSE, recursive = TRUE)
    ggplot2::ggsave(
      filename = protein_plot_file,
      plot = p,
      width = 3.0, height = 4.5, units = "in",
      dpi = 600,
      bg = "white"
    )
  })
```

### Single Protein Logistic Regression Analysis

#### Perform Logistic Regression

```{r}

all_results[["single_protein_logistic_regression"]] <- single_protein_data |>
  dplyr::rename(
    YEARS_SINCE_ONSET = YEARS_SINCE_ONSET.PD
  ) |>
  dplyr::group_by(
    Protein,
  ) |>
  tidyr::nest() |>
  # Set an explicit column ordering
  dplyr::select(
    Protein,
    data,
  ) |>
  #####
  # Create every combination of pair_id and windows
  #####
  tidyr::crossing(window_specs) |>
  furrr::future_pmap(function(Protein, data, window_id, window_start, window_end) {
    # Force future to send this global variable to each worker
    staging_dir <- staging_output_data_dir
    
    logger::log_info("Protein={Protein} window_id={window_id}")
    
    data |>
      #####
      # Restrict our data to the time-slice requested
      #####
      dplyr::filter(
        YEARS_SINCE_ONSET >= window_start,
        YEARS_SINCE_ONSET < window_end,
      ) |>
      #####
      # For participants with more than 1 sample in the window, pick the middle one
      #####
      pick_single_sample_per_participant(strategy = "middle") |>
      #####
      # Save each dataset to a HIVE-style directory structure of parquet files
      #####
      .save_dataset(
        filename = glue::glue("{staging_dir}/single_protein_logistic_regression_data/protein={Protein}/window_id={window_id}/sp_logistic_regression_data.parquet")
      ) |>
      #####
      # Apply the logistic regression model
      #####
      apply_logistic_regression_model_to_window(
        model = stats::as.formula("group ~ SEX + YOB + NPX"),
        exposure = "NPX"
      ) |>
      #####
      # Add info about time-slice requested
      #####
      dplyr::mutate(
        Protein = !!Protein,

        # Reconstruct the window_id using the start and end
        window_id = glue::glue("[{window_start},{window_end})"),
        window_start = window_start,
        window_end = window_end,
      )
    },
    .options = furrr::furrr_options(seed = TRUE)
  ) |>
  dplyr::bind_rows() |>
  #####
  # Ensure the columns are in the same order as previous results
  #####
  dplyr::relocate(
    Protein,
    window_id,
    window_start,
    window_end,
    n_samples,
  )
```

#### Save Results

```{r}

all_results[["single_protein_logistic_regression"]] |>
  readr::write_tsv(
    here::here(staging_output_data_dir, "ukb_single_protein_logistic_regression.tsv.gz")
  )
```

#### Plot Results

##### Balance Measures in Covariates

```{r}

cobalt_long[["single_protein_logistic_regression"]] <- all_results[["single_protein_logistic_regression"]] |>
  dplyr::select(tidyselect::starts_with("cobalt_")) |>
  tidyr::pivot_longer(
    cols = tidyselect::everything(),
    names_to = "covariate",
    values_to = "value"
  ) |>
  dplyr::mutate(
    interval_colours = dplyr::case_when(
      abs(value) >= 0.2 ~ "lightcoral",
      abs(value) >= 0.1 ~ "tan1",
      .default = "palegreen3"
    )
  )

cobalt_plot[["single_protein_logistic_regression"]] <- cobalt_long[["single_protein_logistic_regression"]] |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = value,
      fill = interval_colours
    )
  ) +
  ggplot2::geom_histogram(
    binwidth = 0.025,
    colour = "white",
  ) +
  ggplot2::facet_wrap(
    ~ covariate,
    scales = "free_y",
    labeller = ggplot2::labeller(
      covariate = function(x) gsub("^cobalt_", "", x)
    )
  ) +
  ggplot2::scale_fill_identity() +
  ggplot2::labs(
    x = "Balance Statistic",
    y = "Count"
  ) +
  ggplot2::expand_limits(
    x = c(-1, 1)
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-1, 1, by = 0.2)
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  )

cobalt_hist_plots_pdf[["single_protein_logistic_regression"]] <- here::here(glue::glue("{staging_output_plot_dir}/cobalt_hist_single_protein_logistic_regression.pdf"))
dir.create(dirname(cobalt_hist_plots_pdf[["single_protein_logistic_regression"]]), showWarnings = FALSE, recursive = TRUE)

pdf(cobalt_hist_plots_pdf[["single_protein_logistic_regression"]], width = 15.0, height = 5.0)
cobalt_plot[["single_protein_logistic_regression"]]
dev.off()
```

##### Volcano Plots

```{r}

#####
# Setup base volcano plot
#####
volcano_plot[["single_protein_logistic_regression"]] <- all_results[["single_protein_logistic_regression"]] |>
  plot_volcano(x = "log_or") + ggplot2::geom_point(
    alpha = 0.7,
    shape = 16
  )
```

```{r}
#! label: plot-volcanos

#####
# Plot a single volcano containing all results
#####
ggplot2::ggsave(
  filename = here::here(glue::glue("{staging_output_plot_dir}/volcano_single_protein_logistic_regression.pdf")),
  plot = volcano_plot[["single_protein_logistic_regression"]],
  width = 6.0, height = 4.0, units = "in",
  dpi = 600,
  bg = "white"
)

#####
# Plot a volcano for each window analysed
#####
pdf(here::here(glue::glue("{staging_output_plot_dir}/volcano_by_window_single_protein_logistic_regression.pdf")), width = 6.0, height = 4.0)
for (w in unique(all_results[["single_protein_logistic_regression"]][["window_id"]])) {
  volcano_plot_window <- volcano_plot[["single_protein_logistic_regression"]] +
    ggplot2::geom_point(
      data = dplyr::filter(all_results[["single_protein_logistic_regression"]], window_id == w),
      alpha = 0.7,
      shape = 16
    ) +
    ggplot2::labs(
      subtitle = glue::glue("Window: {w}")
    ) +
    ggplot2::theme_minimal()
  
  print(volcano_plot_window)
}
dev.off()
```

### Protein Ratio Analysis

Subset the data to the proteins of interest:

```{r}

#####
# gene_symbols - columns
#####
gene_symbols[["protein_ratio"]] <- ppoi |>
  tidyr::pivot_longer(
    cols = c(gene_symbol_numerator, gene_symbol_denominator),
    values_to = "gene_symbol"
  ) |>
  dplyr::distinct(
    gene_symbol
  ) |>
  dplyr::arrange(
    gene_symbol
  )
```

#### Calculate Protein Ratios

```{r}
#| label: calc-protein-ratios

ratio_data_ppoi <- ppoi |>
  #####
  # Add on the NPX data for the numerator
  #####
  dplyr::inner_join(
    proteomics_data_long |>
      dplyr::rename(NPX.numerator = NPX),
    by = dplyr::join_by(
      "gene_symbol_numerator" == "Protein"
    ),
    relationship = "many-to-many"
  ) |>
  #####
  # Add on the NPX data for the denominator from the matching sample
  #####
  dplyr::inner_join(
    proteomics_data_long |>
      dplyr::rename(NPX.denominator = NPX),
    by = dplyr::join_by(
      "Sample_ID",
      "gene_symbol_denominator" == "Protein"
    ),
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    ratio = NPX.numerator - NPX.denominator
  )
```

#### Extract Data for Logistic Regression

```{r}
#| label: prep-logistic-regression

lr_data <- ratio_data_ppoi |>
  dplyr::left_join(
    proteomics_data[["row_metadata"]],
    by = dplyr::join_by("Sample_ID")
  ) |>
  dplyr::left_join(
    disease_status,
    by = dplyr::join_by(Participant_ID)
  ) |>
  #####
  # Join participant metadata
  #####
  dplyr::left_join(
    all_participants,
    by = dplyr::join_by(Participant_ID)
  ) |>
  #####
  # Make some variables a factor
  #####
  dplyr::mutate(
    group = factor(group, levels = c("control.eid", "case.eid")),
    SEX = factor(SEX),
    HAS_T2D = factor(HAS_T2D),
    HAS_HYPERTENSION = factor(HAS_HYPERTENSION),
  ) |>
  dplyr::filter(
    !is.na(ratio)
  )
```

#### Perform Logistic Regression

```{r}
#| label: perform-logistic-regression

all_results[["protein_ratio"]] <- lr_data |>
  dplyr::rename(
    YEARS_SINCE_ONSET = YEARS_SINCE_ONSET.PD
  ) |>
  dplyr::group_by(
    pair_id,
    gene_symbol_numerator,
    gene_symbol_denominator,
  ) |>
  tidyr::nest() |>
  # Set an explicit column ordering
  dplyr::select(
    pair_id,
    gene_symbol_numerator,
    gene_symbol_denominator,
    data,
  ) |>
  #####
  # Create every combination of pair_id and windows
  #####
  tidyr::crossing(window_specs) |>
  furrr::future_pmap(function(pair_id, gene_symbol_numerator, gene_symbol_denominator, data, window_id, window_start, window_end) {
    # Force future to send this global variable to each worker
    staging_dir <- staging_output_data_dir
    
    data |>
      #####
      # Restrict our data to the time-slice requested
      #####
      dplyr::filter(
        YEARS_SINCE_ONSET >= window_start,
        YEARS_SINCE_ONSET < window_end,
      ) |>
      #####
      # For participants with more than 1 sample in the window, pick the middle one
      #####
      pick_single_sample_per_participant(strategy = "middle") |>
      #####
      # Save each dataset to a HIVE-style directory structure of parquet files
      #####
      .save_dataset(
        filename = glue::glue("{staging_dir}/logistic_regression_data/pair_id={pair_id}/window_id={window_id}/lr_data.parquet")
      ) |>
      #####
      # Apply the logistic regression model
      #####
      apply_logistic_regression_model_to_window(
        model = stats::as.formula("group ~ SEX + YOB + ratio"),
        exposure = "ratio"
      ) |>
      #####
      # Add info about time-slice requested
      #####
      dplyr::mutate(
        pair_id = pair_id,
        gene_symbol_numerator = gene_symbol_numerator,
        gene_symbol_denominator = gene_symbol_denominator,
        
        # Reconstruct the window_id using the start and end
        window_id = glue::glue("[{window_start},{window_end})"),
        window_start = window_start,
        window_end = window_end,
      )
    },
    .options = furrr::furrr_options(seed = TRUE)
  ) |>
  dplyr::bind_rows() |>
  #####
  # Add back source file info for each PPOI
  #####
  dplyr::left_join(
    ppoi |>
      dplyr::distinct(
        pair_id,
        File,
      ),
    by = dplyr::join_by(pair_id)
  ) |>
  #####
  # Add back assay_id info for each PPOI
  #####
  dplyr::left_join(
    ppoi |>
      dplyr::distinct(
        pair_id,
        gene_symbol_numerator,
        gene_symbol_denominator,
        assay_id_numerator,
        assay_id_denominator,
      ),
    by = dplyr::join_by(
      pair_id,
      gene_symbol_numerator,
      gene_symbol_denominator,
    ),
  ) |>
  #####
  # Ensure the columns are in the same order as previous results
  #####
  dplyr::relocate(
    File,
    pair_id,
    gene_symbol_numerator,
    gene_symbol_denominator,
    assay_id_numerator,
    assay_id_denominator,
    window_id,
    window_start,
    window_end,
    n_samples,
  )
```

#### Save Results

```{r}
#| label: write-results

all_results[["protein_ratio"]] |>
  readr::write_tsv(
    here::here(staging_output_data_dir, "ukb_validation_lists_1-3.tsv.gz")
  )
```

#### Plot Results

##### Balance Measures in Covariates

```{r}
#! label: prep-plot-cobalt-histogram-sex

cobalt_long[["protein_ratio"]] <- all_results[["protein_ratio"]] |>
  dplyr::select(tidyselect::starts_with("cobalt_")) |>
  tidyr::pivot_longer(
    cols = tidyselect::everything(),
    names_to = "covariate",
    values_to = "value"
  ) |>
  dplyr::mutate(
    interval_colours = dplyr::case_when(
      abs(value) >= 0.2 ~ "lightcoral",
      abs(value) >= 0.1 ~ "tan1",
      .default = "palegreen3"
    )
  )

cobalt_plot[["protein_ratio"]] <- cobalt_long[["protein_ratio"]] |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = value,
      fill = interval_colours
    )
  ) +
  ggplot2::geom_histogram(
    binwidth = 0.025,
    colour = "white",
  ) +
  ggplot2::facet_wrap(
    ~ covariate,
    scales = "free_y",
    labeller = ggplot2::labeller(
      covariate = function(x) gsub("^cobalt_", "", x)
    )
  ) +
  ggplot2::scale_fill_identity() +
  ggplot2::labs(
    x = "Balance Statistic",
    y = "Count"
  ) +
  ggplot2::expand_limits(
    x = c(-1, 1)
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-1, 1, by = 0.2)
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  )

cobalt_hist_plots_pdf[["protein_ratio"]] <- here::here(glue::glue("{staging_output_plot_dir}/cobalt_hist_protein_ratio.pdf"))
dir.create(dirname(cobalt_hist_plots_pdf[["protein_ratio"]]), showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = cobalt_hist_plots_pdf[["protein_ratio"]],
  plot = cobalt_plot[["protein_ratio"]],
  width = 15.0, height = 5.0, units = "in",
  dpi = 600,
  bg = "white"
)
```

##### Volcano Plots

```{r}
#! label: prep-plot-volcano

#####
# Setup base volcano plot
#####
volcano_plot[["protein_ratio"]] <- all_results[["protein_ratio"]] |>
  plot_volcano(x = "log_or") + ggplot2::geom_point(
    alpha = 0.7,
    shape = 16
  )
```

```{r}
#! label: plot-volcanos

#####
# Plot a single volcano containing all results
#####
ggplot2::ggsave(
  filename = here::here(glue::glue("{staging_output_plot_dir}/volcano_protein_ratio.pdf")),
  plot = volcano_plot[["protein_ratio"]],
  width = 6.0, height = 4.0, units = "in",
  dpi = 600,
  bg = "white"
)

#####
# Plot a volcano for each window analysed
#####
pdf(here::here(glue::glue("{staging_output_plot_dir}/volcano_by_window_protein_ratio.pdf")), width = 6.0, height = 4.0)
for (w in unique(all_results[["protein_ratio"]][["window_id"]])) {
  volcano_plot_window <- volcano_plot[["protein_ratio"]] +
    ggplot2::geom_point(
      data = dplyr::filter(all_results[["protein_ratio"]], window_id == w),
      alpha = 0.7,
      shape = 16
    ) +
    ggplot2::labs(
      subtitle = glue::glue("Window: {w}")
    ) +
    ggplot2::theme_minimal()
  
  print(volcano_plot_window)
}
dev.off()
```

##### Protein Pairs: NPX and Ratios

```{r}
#! label: plot-pairs-expression-profiles

all_pairs_pdf <- here::here(glue::glue("{staging_output_plot_dir}/all_pairs_UKB.pdf"))
dir.create(dirname(all_pairs_pdf), showWarnings = FALSE, recursive = TRUE)

subdir_pair_id <- here::here(glue::glue("{staging_output_plot_dir}/pair_id"))
dir.create(subdir_pair_id, showWarnings = FALSE, recursive = TRUE)

pdf(all_pairs_pdf, width = 9.0, height = 4.5)
lr_data |>
  dplyr::rename(
    YEARS_SINCE_ONSET = YEARS_SINCE_ONSET.PD
  ) |>
  dplyr::group_by(pair_id) |>
  tidyr::nest() |>
  purrr::pwalk(function(pair_id, data) {
    p <- plot_pair(data, symmetric.y = TRUE)
    
    # Also create individual plot files
    pair_plot_file <- here::here(glue::glue("{staging_output_plot_dir}/pair_id/{pair_id}_UKB.pdf"))
    dir.create(dirname(pair_plot_file), showWarnings = FALSE, recursive = TRUE)
    ggplot2::ggsave(
      filename = pair_plot_file,
      plot = p,
      width = 9.0, height = 4.5, units = "in",
      dpi = 600,
      bg = "white"
    )
  })
dev.off()
```

## Upload Outputs to UKB-RAP

```{r}
#! label: create-tarball

# Prepare tarball
stage_dir <- glue::glue(getwd(), "/staging/")
tarball_base <- staging_parent_dir |> stringr::str_replace("/([^\\/]+)$", "_\\1") |> stringr::str_remove(glue::glue(getwd(), "/staging/"))
staging_parent_subdir <- staging_parent_dir |> stringr::str_remove(glue::glue(getwd(), "/staging/"))

system(
  glue::glue('cd {stage_dir} && tar -zcf {tarball_base}.tar.gz {staging_parent_subdir}'),
  intern = TRUE,
  ignore.stderr = TRUE
)
```

```{r}
#! label: upload-tarball

# Upload to UKB-RAP
local_file <- glue::glue('staging/{tarball_base}.tar.gz')
rap_path <- glue::glue("{ukb_rap_working_dir}/protein_expression_analysis/{filename}.tar.gz", filename = staging_parent_dir |> stringr::str_replace("/([^\\/]+)$", "_\\1") |> stringr::str_remove(glue::glue(getwd(), "/staging/")))

ukb_rap_upload(
  from = local_file,
  rap_path = rap_path
)
```
