---
title: "Cohort F QC Report with Matching Sample"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
# install.packages(c("purrr", "htmltools", "plotly", "irlba", "Rtsne", "gtsummary", "mt", "future", "MatchIt", "relaimpo"))

library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(broom)
library(uwot)
library(mt)
library(irlba)
library(Rtsne)
library(gtsummary)
library(future)
library(MatchIt)
library(knitr)

source("/home/rstudio/gnpc/scope-analysis/R/qc_clinical.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/qc_proteomics.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_pca.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_tsne.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_umap.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_scatter.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_boxplot.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_plsda.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_variance_distribution.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_lod.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/linear_regression.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/tbl_coef_compare.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/qc_summary.R", echo = FALSE, print.eval = FALSE)
source("/home/rstudio/gnpc/scope-analysis/R/plot_variance_stacked.R", echo = FALSE, print.eval = FALSE)
```

## Load Data and Selected Metadata Columns
```{r, include=TRUE}
input_version <- "v6"
selected_contributor_code <- "F"

metadata <- readRDS(paste0("/home/rstudio/data_prep/metadata/clinical_metadata_tbl_", input_version, "_contributor_", selected_contributor_code, ".rds"))
annotations <- readRDS(paste0("/home/rstudio/data_prep/annotation/annotation_tbl_", input_version, ".rds"))
protein_intensity <- readRDS(paste0("/home/rstudio/data_prep/proteomics/proteomic_tbl_", input_version, "_contributor_", selected_contributor_code, ".rds"))
```

## Table 1 Descriptive statistics for continuous and categorical variables (Before PCA outliers filtering and matching)
```{r, summary-1, include=TRUE}
summary_cols <- c(
  "pd", "ad", "ftd", "als", "race", "age_at_visit", "sex", "years_of_education",
  "resting_heart_rate_pulse",
  "systolic_blood_pressure_sitting", "diastolic_blood_pressure_sitting",
  "alcohol_hx", "smoking_hx", "bmi_cat", "diabetes", "mci_sci", "hyperlipidaemia",
  "depression", "cognitive_test_score",
  "cognitive_test_battery", "depression_test_battery", "sample_matrix", "sequential_visit_number"
)
continuous_cols <- c(
  "age_at_visit", "years_of_education", "resting_heart_rate_pulse",
  "systolic_blood_pressure_sitting", "diastolic_blood_pressure_sitting",
  "cognitive_test_score"
)

metadata_w_dummie_col <- metadata
metadata_w_dummie_col$Summary <- "summary"
create_summary_table(
  data = metadata_w_dummie_col,
  id_col = "person_id",
  group_col = "Summary",
  summary_cols = summary_cols,
  continuous_cols = continuous_cols,
  statistic_type = "comprehensive"
)
```



## Table 2 Descriptive statistics for continuous and categorical variables per PD status (Before PCA outliers filtering and matching)
```{r, include=TRUE}
create_summary_table(
  data = metadata,
  id_col = "person_id",
  group_col = "pd",
  summary_cols = summary_cols,
  continuous_cols = continuous_cols,
  statistic_type = "comprehensive",
  show_pvalues = TRUE,
  continuous_val_test_type = "auto"
)
```


## Table 3 Descriptive statistics for continuous and categorical variables (Before PCA outliers filtering and matching) by sample
```{r, summary-3, include=TRUE}
metadata_w_dummie_col <- metadata
metadata_w_dummie_col$Summary <- "summary"
create_summary_table(
  data = metadata_w_dummie_col,
  id_col = "sample_id",
  group_col = "Summary",
  summary_cols = summary_cols,
  continuous_cols = continuous_cols,
  statistic_type = "comprehensive"
)
```


## Table 4 Descriptive statistics for continuous and categorical variables per PD status (Before PCA outliers filtering and matching) by sample
```{r, include=TRUE}
create_summary_table(
  data = metadata,
  id_col = "sample_id",
  group_col = "pd",
  summary_cols = summary_cols,
  continuous_cols = continuous_cols,
  statistic_type = "comprehensive",
  show_pvalues = TRUE,
  continuous_val_test_type = "auto"
)
```


## Protein intensity transformed to log2 scale
```{r Protein intensity log2 transformation, results='asis'}
protein_intensity_log2 <- protein_intensity |> mutate(across(where(is.numeric), ~ log2(.x + 1)))
```


## Select a set of more relevant columns
```{r first-filtering, include=TRUE}
## select metadata columns
selected_metadata_cols <- c(
  "sample_id", "person_id", "visit", "contributor_code", "sequential_visit_number",
  "age_at_visit", "sex", "race", "years_of_education",
  "resting_heart_rate_pulse", "systolic_blood_pressure_sitting",
  "diastolic_blood_pressure_sitting", "alcohol_hx", "smoking_hx", "total_years_smoked",
  "bmi_cat", "pd", "ad", "ftd", "als", "stroke", "tbi", "mci_sci", "diabetes",
  "hypertension", "hyperlipidaemia", "depression", "anxiety",
  "depression_inventory_score", "depression_test_battery", "cdr", "cognitive_test_date",
  "cognitive_test_score", "computed_cognitive_test_score", "cognitive_test_battery",
  "computed_clinical_diagnosis", "computed_cognitive_impairment", "sample_matrix"
)

filtered_metadata <- metadata |>
  select(all_of(selected_metadata_cols))
```

## Missingness in Cohort Clinical Data
```{r, include=TRUE}
# Identify variable types
## Note that at this step, it started with unfiltered metadata
numeric_vars <- metadata |>
  select(where(is.numeric)) |>
  names()
categorical_vars <- metadata |>
  select(where(~ is.character(.) | is.factor(.))) |>
  names()

# Missingness summary for numeric variables
missing_numeric <- metadata |>
  summarise(across(all_of(numeric_vars), ~ mean(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_pct"
  ) |>
  mutate(type = "numeric")

# Missingness summary for categorical variables
missing_categorical <- metadata |>
  summarise(across(all_of(categorical_vars), ~ mean(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_pct"
  ) |>
  mutate(type = "categorical")

# Combine both summaries
missing_summary <- bind_rows(missing_numeric, missing_categorical)

# View
ggplot(
  missing_summary |> filter(missing_pct > 0.0),
  aes(x = reorder(variable, missing_pct), y = missing_pct, fill = type)
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Column Missingness by Variable Type",
    x = "Variable",
    y = "Missingness (%)"
  ) +
  theme_minimal()
```

Then filter columns with more than 10% missingness, as well as all "-1" value columns. These columns may contain other values from different contributors but it may be missing (NA) for certain contributors. 
```{r, include=TRUE}
## more than 10% missingness cols are filtered except for required ones : age, sex, AD/FTD/ALS, bmi, alcohol and smoking history
required_variables <- c("age_at_visit", "sex", "ad", "ftd", "als", "race", "years_of_education", "resting_heart_rate_pulse", "systolic_blood_pressure_sitting", "diastolic_blood_pressure_sitting", "alcohol_hx", "smoking_hx", "bmi_cat", "mci_sci", "cognitive_test_score", "depression_test_battery")

missing_pct_threshold <- 0.10
columns_with_high_missingness <- missing_summary |>
  filter(missing_pct >= missing_pct_threshold) |>
  pull(variable)

## remove these high missingness, as well as all "-1" (meaning NA) value columns
filtered_metadata <- metadata |>
  select(-all_of(columns_with_high_missingness), all_of(required_variables))

## remove these all "-1" (meaning NA) value columns
columns_with_all_neg1s <- filtered_metadata |>
  select(where(~ all(.x == -1, na.rm = TRUE))) |>
  colnames()
filtered_metadata <- filtered_metadata |>
  select(-all_of(columns_with_all_neg1s))
```


## Filter for the latest sample per patient
```{r longitudinal-last-sample-filtering, include=TRUE}
# filter for the latest sample per patient
## Define a util function to get top_n visits by age per patient
get_top_n_samples_per_patient <- function(df, top_n = 1, order_col = age_at_visit, desc = TRUE) {
  df |>
    group_by(person_id) |>
    {
      if (desc) {
        arrange(., dplyr::desc({{ order_col }}), .by_group = TRUE)
      } else {
        arrange(., {{ order_col }}, .by_group = TRUE)
      }
    } |>
    slice_head(n = top_n) |>
    ungroup() |>
    pull(sample_id)
}
top_n <- 1
top_n_age_at_visit_sample_ids <- get_top_n_samples_per_patient(filtered_metadata,
  top_n = top_n,
  order_col = "sequential_visit_number", desc = TRUE
)

longti_filtered_metadata <- filtered_metadata |> filter(sample_id %in% top_n_age_at_visit_sample_ids)
longti_filtered_metadata_sample_ids <- longti_filtered_metadata |> pull(sample_id)
protein_intensity_log2 <- protein_intensity_log2 |> filter(sample_id %in% longti_filtered_metadata_sample_ids)
```




## Protein intensity sample median/PCA outliers
```{r protein-intensity-sample-median}
# Define protein list (columns to include)
protein_list <- names(select(protein_intensity_log2, -sample_id))

# Calculate per-sample median
sample_median <- apply(protein_intensity_log2 |> dplyr::select(all_of(protein_list)), 1, median)

# Compute mean and SD of sample medians
mean_protein <- mean(sample_median)
sd_protein <- sd(sample_median)

# Flag outliers beyond ±5 SD
plot_data <- data.frame(
  sample_id = protein_intensity_log2$sample_id,
  sample_median = sample_median,
  outlier_sd = sample_median > mean_protein + 5 * sd_protein |
    sample_median < mean_protein - 5 * sd_protein
)

# count the result rows
nrow(plot_data)
# count the result for outliers
nrow(plot_data |> filter(outlier_sd == TRUE))
print(plot_data |> filter(outlier_sd == TRUE))
```


```{r pca-outlier-filtering, include=TRUE}
# Detect (some) PCA outliers
pca_outlier_row_numbers <- mt::pca.outlier(protein_intensity_log2 |> select(-sample_id))$outlier
pca_outlier_sample_ids <- protein_intensity_log2 |>
  slice(pca_outlier_row_numbers) |>
  pull(sample_id)
cat("Number of PCA outliers :", length(pca_outlier_sample_ids), "\n")
filtered_protein_intensity_log2 <- protein_intensity_log2 |> filter(!sample_id %in% pca_outlier_sample_ids)
filtered_metadata <- longti_filtered_metadata |> filter(!sample_id %in% pca_outlier_sample_ids)
```
** Test if protein intensity sample median outliers are within the PCA outliers
```{r filtering-outliers, include=TRUE}
all(plot_data |> filter(outlier_sd == TRUE) |> pull(sample_id) %in% pca_outlier_sample_ids)

## Check if these sample median outliers are also in the PCA outliers, if not, filter those out from further analysis
if (!all(plot_data |> filter(outlier_sd == TRUE) |> pull(sample_id) %in% pca_outlier_sample_ids)) {
  # filtering the sample median outliers
  # this is done only for cohort F as these sample median outliers are not included in the PCA outliers
  sample_median_outlier_ids <- plot_data |>
    filter(outlier_sd == TRUE) |>
    pull(sample_id)
  filtered_metadata <- filtered_metadata |> filter(!sample_id %in% sample_median_outlier_ids)
  filtered_protein_intensity_log2 <- filtered_protein_intensity_log2 |> filter(!sample_id %in% sample_median_outlier_ids)

  # how many samples left
  cat("Number of samples remain for protein ratio validation :", nrow(filtered_metadata), "\n")
}
```


## PCA after Outliers Removed
```{r add-dummie-ID-cols-to-non-outliers}
copied_filtered_protein_intensity_log2 <- filtered_protein_intensity_log2
copied_filtered_protein_intensity_log2$ID <- copied_filtered_protein_intensity_log2$sample_id
copied_filtered_protein_intensity_log2$sample_id <- NULL

copied_filtered_metadata <- filtered_metadata
copied_filtered_metadata$ID <- copied_filtered_metadata$sample_id
copied_filtered_metadata$sample_id <- NULL
```
```{r non-outliers-pca-panels}
selected_categorical_vars <- categorical_vars[
  vapply(categorical_vars, function(cat_var) {
    n_unique <- length(unique(copied_filtered_metadata[[cat_var]]))
    n_unique <= 50 && n_unique > 1
  }, logical(1))
]
pca_res <- plot_PCA(
  protein_intensity = copied_filtered_protein_intensity_log2,
  metadata = copied_filtered_metadata,
  var_name = selected_categorical_vars,
  IDs_for_outliers = NULL, # c("ID")
  nsd = 6,
  exact = TRUE,
  point_size = 1
)
pca_res$plots
```

## tSNE after Outliers Removed

```{r non-outliers-tsne, results='asis'}
for (cat_var in categorical_vars) {
  ## do not plot for vars with too many or too few values
  if (length(unique(copied_filtered_metadata[[cat_var]])) <= 50 &
    length(unique(copied_filtered_metadata[[cat_var]])) > 1) {
    tsne_plot <- plot_tSNE(
      protein_intensity = copied_filtered_protein_intensity_log2,
      metadata = copied_filtered_metadata,
      var_name = cat_var,
      IDs_for_tooltip = c("ID"),
      title_name = paste0("tSNE Plot (", cat_var, ")"),
      perplexity = 10,
      point_size = 1
    )
    print(tsne_plot)
  }
}
```

## UMAP after Outliers Removed
```{r non-outliers-umap, results='asis'}
for (cat_var in categorical_vars) {
  ## do not plot for vars with too many or too few values
  if (length(unique(copied_filtered_metadata[[cat_var]])) <= 50 &
    length(unique(copied_filtered_metadata[[cat_var]])) > 1) {
    umap_plot <- plot_UMAP(
      protein_intensity = copied_filtered_protein_intensity_log2,
      metadata = copied_filtered_metadata,
      var_name = cat_var,
      IDs_for_tooltip = c("ID"),
      title_name = paste0("UMAP Plot (", cat_var, ")"),
      point_size = 1
    )
    print(umap_plot)
  }
}
```


## Variance Distribution
```{r variance_plot_table, results='asis'}
## plot_variance_distribution will by default perform log10 transformation on the protein intensity
## so input here is non-log transformed
result <- plot_variance_distribution(protein_intensity |> filter(!sample_id %in% pca_outlier_sample_ids))
print(result$plot)

# Text and spacing
cat("\n\n###    \n\n")

DT::datatable(
  result$variance_table,
  extensions = "Buttons",
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel", "pdf"),
    pageLength = 10,
    scrollX = TRUE
  ),
  filter = "top",
  rownames = FALSE
)
```



## Matching Samples
```{r matching-function}
match_stratified_cohort <- function(
  data,
  predictors = c("age_at_visit", "sex", "ad"),
  response = "pd",
  tolerance_in_age = 5,
  ratio = 1,
  seed = 42,
  method = "nearest",
  distance = "mahalanobis",
  formula = "pd ~ age_at_visit + sex + ad"
) {
  # Check required columns
  required_vars <- all(all.vars(formula) %in% names(data))
  if (!required_vars) {
    stop("Some variables in the formula are not found in the data.")
  }

  # Build formula for sample matching
  formula <- reformulate(predictors, response = response)

  # Perform Mahalanobis nearest neighbor matching
  set.seed(seed)
  m.out <- matchit(
    formula,
    data = data,
    method = method,
    distance = distance,
    caliper = c(age_at_visit = tolerance_in_age),
    ratio = ratio
  )
  matched_metadata <- match.data(m.out)

  return(
    list(
      m.out = m.out,
      matched_metadata = matched_metadata
    )
  )
}
```

```{r matching-samples, include=TRUE}
matching_ratio <- 2 # ratio 2 for cohort F
matched_output <- match_stratified_cohort(
  data = filtered_metadata,
  predictors = c("age_at_visit", "sex", "ad"), ## note here the predictors and response is for sample matching
  response = "pd",
  tolerance_in_age = 5,
  ratio = matching_ratio,
  seed = 42,
  method = "nearest",
  distance = "mahalanobis"
)
# Check covariate balance before and after
summary(matched_output$m.out)
# Run regression on matched dataset
validate_lm <- glm(pd ~ age_at_visit + sex, data = matched_output$matched_metadata, family = binomial())
summary(validate_lm)

matched_sample_ids <- matched_output$matched_metadata |> pull(sample_id)
cat("Number of matched samples :", length(matched_sample_ids), "\n")
matched_filtered_protein_intensity_log2 <- filtered_protein_intensity_log2 |> filter(sample_id %in% matched_sample_ids)
matched_filtered_metadata <- matched_output$matched_metadata
```

## Table 5 Descriptive statistics for continuous and categorical variables (After PCA outliers filtering and matching)
```{r, pca-outlier-rm-summary-1, include=TRUE}
summary_cols <- c(
  "pd", "ad", "ftd", "als", "race", "age_at_visit", "sex", "years_of_education",
  "resting_heart_rate_pulse",
  "systolic_blood_pressure_sitting", "diastolic_blood_pressure_sitting",
  "alcohol_hx", "smoking_hx", "bmi_cat", "diabetes", "mci_sci", "hyperlipidaemia",
  "depression", "cognitive_test_score",
  "cognitive_test_battery", "depression_test_battery", "sample_matrix", "sequential_visit_number"
)
continuous_cols <- c(
  "age_at_visit", "years_of_education", "resting_heart_rate_pulse",
  "systolic_blood_pressure_sitting", "diastolic_blood_pressure_sitting",
  "cognitive_test_score"
)

summary_cols <- summary_cols[!summary_cols %in% columns_with_all_neg1s]
continuous_cols <- continuous_cols[!continuous_cols %in% columns_with_all_neg1s]

matched_filtered_metadata_w_dummie_col <- matched_filtered_metadata
matched_filtered_metadata_w_dummie_col$Summary <- "summary"
create_summary_table(
  data = matched_filtered_metadata_w_dummie_col,
  id_col = "person_id",
  group_col = "Summary",
  summary_cols = summary_cols,
  continuous_cols = continuous_cols,
  statistic_type = "comprehensive"
)
```


## Table 6 Descriptive statistics for continuous and categorical variables per PD status (After PCA outliers filtering and matching)
```{r, include=TRUE}
create_summary_table(
  data = matched_filtered_metadata,
  id_col = "person_id",
  group_col = "pd",
  summary_cols = summary_cols,
  continuous_cols = continuous_cols,
  statistic_type = "comprehensive",
  show_pvalues = TRUE,
  continuous_val_test_type = "auto"
)
```


#### Matching Samples PCA Panels
```{r add-dummie-ID-cols-matching-samples}
copied_matched_filtered_protein_intensity_log2 <- matched_filtered_protein_intensity_log2
copied_matched_filtered_protein_intensity_log2$ID <- copied_matched_filtered_protein_intensity_log2$sample_id
copied_matched_filtered_protein_intensity_log2$sample_id <- NULL

copied_matched_filtered_metadata <- matched_filtered_metadata
copied_matched_filtered_metadata <- copied_matched_filtered_metadata |> mutate(age_over_50 = if_else(age_at_visit > 50, "Yes", "No"))
copied_matched_filtered_metadata$ID <- copied_matched_filtered_metadata$sample_id
copied_matched_filtered_metadata$sample_id <- NULL
```
```{r pca-panels}
selected_cat_vars_for_matching_samples <- c("sex", "age_over_50", "pd", "ad")

pca_res <- plot_PCA(
  protein_intensity = copied_matched_filtered_protein_intensity_log2,
  metadata = copied_matched_filtered_metadata,
  var_name = selected_cat_vars_for_matching_samples,
  title_name = "PCA Plot",
  nsd = 6,
  exact = TRUE,
  IDs_for_outliers = NULL,
  ellipse_conf = 0.95,
  point_size = 1,
  label_outliers = TRUE,
  custom_colors = NULL
)
pca_res$plots
```


## Matching Samples tSNE
```{r tsne-matching-samples, results='asis'}
for (cat_var in selected_cat_vars_for_matching_samples) {
  tsne_plot <- plot_tSNE(
    prot = copied_matched_filtered_protein_intensity_log2,
    metadata = copied_matched_filtered_metadata,
    var_name = cat_var,
    IDs_for_tooltip = c("ID"),
    title_name = paste0("tSNE Plot (", cat_var, ")"),
    perplexity = 10,
    point_size = 1
  )
  print(tsne_plot)
}
```


## Matching Samples UMAP
```{r umap-matching-samples, results='asis'}
for (cat_var in selected_cat_vars_for_matching_samples) {
  umap_plot <- plot_UMAP(
    protein_intensity = copied_matched_filtered_protein_intensity_log2,
    metadata = copied_matched_filtered_metadata,
    var_name = cat_var,
    IDs_for_tooltip = c("ID"),
    title_name = paste0("UMAP Plot (", cat_var, ")"),
    point_size = 1
  )
  print(umap_plot)
}
```
